
\begin{abstract}
  大语言模型在问答、对话生成、内容创作和辅助决策等任务中表现出较强的语言理解与生成能力。然而，当输入或输出涉及性别、种族、宗教、政治立场等敏感属性时，模型仍可能生成带有刻板印象、差别对待或隐性歧视色彩的不公平回复。与此同时，当前主流大语言模型多以 API 或封闭服务形式部署，研究者通常无法获取模型参数、训练数据和内部表征，导致许多依赖白盒条件的偏见缓解方法难以直接应用于真实场景。针对上述问题，本文面向黑盒大语言模型输出场景，研究基于反事实检测与约束重写的偏见修正方法，构建由偏见识别、风险验证、定向重写和一致性校验组成的输出级后处理框架。

  针对黑盒条件下敏感属性表达复杂、偏见信号难以稳定识别以及单次判断容易误判的问题，本文提出一种基于反事实构造与二阶段验证的偏见识别方法。该方法首先从自然语言输入中识别敏感属性信息，并构造语义主干保持一致、仅在敏感属性维度发生替换的反事实 prompt；然后分别生成原始回复和反事实回复，从语义偏移、态度变化和生成偏好差异三个维度刻画输出对敏感属性变化的响应差异，得到候选偏见分数；在此基础上，进一步围绕回复差异命题设计融合公平性与事实性的二阶段验证机制，对候选偏见样本进行高置信筛选与确认。针对已识别偏见回复中重写范围难以控制、整体改写容易造成信息丢失和语义漂移的问题，本文进一步提出一种基于高风险片段定位与最小编辑约束的偏见重写方法。该方法首先对偏见回复进行句子级、短语级和 span 级风险定位，并结合片段偏见风险、重写代价和综合优先级确定最终重写单元；随后在内容保持、长度与结构控制等约束下执行局部定向重写；最后通过信息保留、语义连贯性和新风险引入检测等一致性校验机制，实现对重写结果的接受与回退控制。

  在实验部分，本文按照“先识别、后重写”的流程，在 BOLD 和 RedditBias 数据集上构造原始输入与反事实输入对，对所提方法进行系统验证。偏见识别阶段以 HONEST、TWBias 等方法作为对比，采用情感差值、毒性差值和外部评审一致率等指标进行评估。实验结果表明，本文方法在偏见识别阶段的平均情感差值、平均毒性差值以及外部评审一致率分别达到 \textbf{[待填写]}、\textbf{[待填写]} 和 \textbf{[待填写]}；引入二阶段验证后，高置信偏见样本的筛选质量进一步提升，相关指标较未使用验证机制时提高 \textbf{[待填写]}。偏见重写阶段以原始回复、Self-Debiasing 和 Moral Self-Correction 为对比方法，从重写后情感差值、毒性差值和编辑距离比例等方面进行评价。实验结果表明，本文方法在偏见重写阶段可将重写后平均情感差值和平均毒性差值分别降低至 \textbf{[待填写]} 和 \textbf{[待填写]}，并将平均编辑距离比例控制在 \textbf{[待填写]}，说明该方法能够在减弱偏见表达的同时较好保持原始回复的主要信息与整体可用性。

  本文在不修改目标模型参数和训练过程的前提下，实现了面向黑盒大语言模型输出的偏见识别与可控校正，为大语言模型输出公平性治理提供了一种具有可解释性、可操作性和一定通用性的技术方案。
\end{abstract}


\begin{abstractEn}
  Large language models have demonstrated strong capabilities in question answering, dialogue generation, content creation, and decision support. However, when prompts or responses involve sensitive attributes such as gender, race, religion, and political ideology, these models may still produce unfair outputs containing stereotypes, disparate treatment, or implicit discrimination. Meanwhile, most mainstream large language models are deployed as black-box services through APIs, making their parameters, training data, and internal representations inaccessible. As a result, many existing debiasing methods that rely on white-box access are difficult to apply in real deployment settings. To address this problem, this thesis studies a bias correction approach for black-box large language model outputs based on counterfactual detection and constrained rewriting, and builds an output-level post-processing framework composed of bias identification, risk verification, targeted rewriting, and consistency checking.

  To address the challenges of complex sensitive-attribute expressions, unstable bias signals, and frequent misjudgments under black-box conditions, this thesis proposes a bias identification method based on counterfactual construction and two-stage verification. The method first identifies sensitive attributes from natural language inputs and constructs counterfactual prompts that preserve the core semantics while replacing only the sensitive attributes. It then generates original responses and counterfactual responses, and measures their differences from three perspectives: semantic shift, attitude change, and generation preference divergence, thereby obtaining candidate bias scores. On this basis, a two-stage verification mechanism integrating fairness and factuality is further designed to extract difference propositions and perform high-confidence verification of candidate biased samples. To address the difficulty of controlling rewrite scope and the semantic drift caused by holistic rewriting, this thesis further proposes a bias rewriting method based on high-risk span localization and minimal-edit constraints. The method first performs sentence-level, phrase-level, and span-level risk localization, then determines final rewrite units according to bias risk, rewrite cost, and integrated priority, and finally conducts local targeted rewriting under constraints on content preservation, length, and structure. A consistency-checking mechanism is further introduced to evaluate information preservation, semantic coherence, and newly introduced risks, so as to decide whether the rewritten result should be accepted or rolled back.

  Experiments are conducted in a pipeline of "identify first, then rewrite" on BOLD and RedditBias, where original prompts and counterfactual prompts are constructed in pairs. In the bias identification stage, the proposed method is compared with baselines such as HONEST and TWBias, and is evaluated by sentiment difference, toxicity difference, and agreement with external judges. Results show that the proposed method achieves \textbf{[to be filled]}, \textbf{[to be filled]}, and \textbf{[to be filled]} on average sentiment difference, average toxicity difference, and judge agreement, respectively. After introducing the two-stage verification mechanism, the quality of high-confidence biased sample selection is further improved by \textbf{[to be filled]}. In the bias rewriting stage, the proposed method is compared with the original response, Self-Debiasing, and Moral Self-Correction. It reduces the rewritten sentiment difference and toxicity difference to \textbf{[to be filled]} and \textbf{[to be filled]}, respectively, while keeping the average edit ratio at \textbf{[to be filled]}. These results indicate that the proposed method can effectively mitigate biased expressions while preserving the major information and usability of the original response.

  Without modifying model parameters or the training process, this thesis achieves effective bias identification and controllable correction for black-box large language model outputs, and provides an interpretable, practical, and relatively general technical solution for output fairness governance of large language models.
\end{abstractEn}
